# Working with Files

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the tradeoffs between scripts and notebooks
* Save code in R scripts
* Get or set the R working directory
* Identify RDS, CSV, TSV files and functions for reading these
* Inspect the structure of a data frame
:::

This chapter will help you get started working with files in R. First, we
explain how to save (and load) your code, so that you can work on it
incrementally and share it with colleagues. After that, we explain how to read
a dataset from a file into R.


## Saving & Loading Code

:::{.callout-tip}
When you start a new project, it's a good idea to create a specific directory
for all of the project's files. If you're using R, you should also store your R
code in that directory. As you work, periodically save your code.
:::

Most of the time, you won't just write code directly into the R console.
Reproducibility and reusability are important benefits of R over
point-and-click software, and in order to realize these, you have to save your
code to your computer's hard drive.

The most common way to save R code is as an R **script** with the extension
`.R` (see @sec-reading-files for more about extensions). Editing a script is
similar to editing any other text file. You can write, delete, copy, cut, and
paste code.

In RStudio, you can create a new R script with this menu option:

```{.none}
File -> New File -> R Script
```

This will open a new pane in RStudio, like this:

![
  How RStudio typically looks after opening a new R Script.
](/images/rstudio_script.png)

The new pane is the scripts pane, which displays all of the R scripts you're
editing. Each script appears in a separate tab. In the screenshot, only one
script, the new script, is open.

Every line in an R script must be valid R code. Anything else you want to write
in the script (notes, documentation, etc.) must be placed in a comment.

Arrange your code in the order of the steps to solve the problem, even if you
write some parts before others. Comment out or delete any lines of code that
you try but ultimately decide you don't need. Make sure to save the file
periodically so that you don't lose your work. Following these guidelines will
help you stay organized and make it easier to share your code with others
later.

:::{.callout-tip}
While editing, you can run the current line in the R console by pressing
`Ctrl`-`Enter` on Windows and Linux, or `Cmd`-`Enter` on macOS. This way you
can test and correct your code as you write it.
:::


### Running Scripts

You can **source** (that is, run) an entire R script by calling the `source`
function with the path to the script as an argument. This is also what the
"Source on Save" check box refers to in RStudio. The code runs in order, only
stopping if an error occurs.

For instance, if you save the script as `my_cool_script.R`, then you can enter
`source("my_cool_script.R")` in the console to run the entire script. Pay
attention to the path---it may be different on your computer.


### Notebooks: Quarto & R Markdown

In the context of data science, a **notebook** is an interactive file that can
store a mix of code, formatted text, and images. With a notebook, you can
write, run code, and view results all in one place. Viewing and editing a
notebook requires a web browser or IDE. Some notebooks can also be converted to
static documents, such as PDFs. Comments are a good way keep notes as you
develop and run your code, but notebooks provide much more flexibility of
expression. Notebooks are a kind of [literate programming][lit].

[lit]: https://en.wikipedia.org/wiki/Literate_programming

Notebooks excel when you want to do highly interactive work and/or want to
communicate results. Use notebooks to prototype code, analyze data, refine
plots, generate documents and presentations, and practice programming. Scripts
excel when you want to reuse code (and perhaps share it as a package) or want
to run code that doesn't require much user interaction (such as time-consuming
computations you'll run on a server or high-performance computing cluster). The
remainder of this reader assumes you're using an R script rather than the R
console or a notebook, unless otherwise noted.

[Quarto][] is a popular notebook format and system for R. It also supports
Python, Julia, and other programming languages. Quarto files have the extension
`.qmd`. Quarto is based on an older notebook format, [R Markdown][rmd], which
only supports R and can't be converted to as many kinds of documents. R
Markdown is still widely used. R Markdown files have the extension `.Rmd`.

[Quarto]: https://quarto.org/
[rmd]: https://rmarkdown.rstudio.com/

:::{.callout-important}
In order to use [Quarto][], you must first download and install it. It is not
included with R or RStudio.
:::

After installing Quarto, you can create a new Quarto file in RStudio with this
menu option:

```{.none}
File -> New -> Quarto Document...
```

RStudio will prompt you to provide some details about the purpose of the file.

Notebooks are subdivided into **chunks** (or cells). You can create as many
chunks as you like, but each chunk can contain only one kind of content. You
can run a code chunk by clicking on the chunk and pressing `Ctrl`-`Enter`. The
notebook will display the result.

Markdown is a simple language you can use to add formatting to (non-code) text
in a notebook. For example, surrounding a word with asterisks, as in `Let
*sleeping* dogs lie`, makes the surrounded word italic. You can find a short,
interactive tutorial about Markdown [here][mdtutorial].

[mdtutorial]: https://www.markdowntutorial.com/

:::{.callout-note title="See Also"}
To learn more about Quarto, see the [official documentation][quarto-docs].

[quarto-docs]: https://quarto.org/docs/guide/
:::

<!--
Now that you know how file systems and paths work, you're ready to learn how to
save your R code to a file. R code is usually saved into an **R script**
(extension `.R`) or an **R Markdown file** (extension `.Rmd`). R scripts are
slightly simpler, so let's focus on those.

Editing a script is similar to editing any other text document. You can write,
delete, copy, cut, and paste text. You can also save the file to your
computer's file system. When you do, pay attention to where you save the file,
as you might need it later.
-->


## The Working Directory

The **working directory** is the starting point R uses for relative paths (see
@sec-file-systems for a refresher on file paths). Think of the working
directory as the directory R is currently "at" or watching.

The function `getwd` returns the absolute path for the current working
directory, as a string. It doesn't require any arguments:

```{r}
getwd()
```

On your computer, the output from `getwd` will likely be different. This is a
very useful function for getting your bearings when you write relative paths.
If you write a relative path and it doesn't work as expected, the first thing
to do is check the working directory.

The related `setwd` function changes the working directory. It takes one
argument: a path to the new working directory. Here's an example:

```{r}
#| eval: false
setwd("..")

# Now check the working directory.
getwd()
```

:::{.callout-warning}
In your R scripts and notebooks, avoid calls to `setwd`. They make your code
more difficult to understand and to run on other computers. Use appropriate
relative paths instead.

In the R console, it's okay to occasionally use `setwd`. You might need to
change the working directory before you run some code. R's default working
directory is your home directory. In some cases, such as when you open a
project, RStudio will automatically change the working directory. However, it
doesn't always change the working directory, so `setwd` is sometimes still
necessary.
:::

Another function that's useful for dealing with the working directory and file
system is `list.files`. The `list.files` function returns the names of all of
the files and directories inside of a directory. It accepts a path to a
directory as an argument, or assumes the working directory if you don't pass a
path. For instance:

```{r}
# List files and directories in /home/.
list.files("/home/")

# List files and directories in the working directory.
list.files()
```

As usual, since you have a different computer, you're likely to see different
output if you run this code. If you call `list.files` with an invalid path or
an empty directory, the output is `character(0)`:

```{r}
list.files("/this/path/is/fake/")
```

Later on, you'll learn about what `character(0)` means more generally.




## Reading Files {#sec-reading-files}

Analyzing datasets is one of the most common things to do in R. The first step
is to get R to read your data. Datasets come in a variety of file formats, and
you need to identify the format in order to tell R how to read the data.

Most of the time, you can guess the format of a file by looking at its
**extension**, the characters (usually three) after the last dot `.` in the
filename. For example, the extension `.jpg` or `.jpeg` indicates a [JPEG image
file][jpg]. Some operating systems hide extensions by default, but you can find
instructions to change this setting online by searching for "show file
extensions" and your operating system's name. The extension is just part of the
file's name, so it should be taken as a hint about the file's format rather
than a guarantee.

[jpg]: https://en.wikipedia.org/wiki/JPEG

R has built-in functions for reading a variety of formats. The R community also
provides packages (@sec-packages) to read even more formats. For now, let's
focus on datasets that can be read with R's built-in functions.

Here are several formats that are frequently used to distribute data, along
with the name of a built-in function or contributed package that can read the
format:

| Name                        | Extension  | Function or Package | Tabular?  | Text?
| :-------------------------- | :--------  | :------------------ | :-------- | :----
| Comma-separated Values      | `.csv`     | `read.csv`          | Yes       | Yes
| Tab-separated Values        | `.tsv`     | `read.delim`        | Yes       | Yes
| Fixed-width File            | `.fwf`     | `read.fwf`          | Yes       | Yes
| Microsoft Excel             | `.xlsx`    | readr package       | Yes       | No
| Microsoft Excel 1993-2007   | `.xls`     | readr package       | Yes       | No
| [Apache Arrow][arrow]       | `.feather` | arrow package       | Yes       | No
| R Data                      | `.rds`     | `readRDS`           | Sometimes | No
| R Data                      | `.rda`     | `load`              | Sometimes | No
| Plaintext                   | `.txt`     | `readLines`         | Sometimes | Yes
| Extensible Markup Language  | `.xml`     | xml2 package        | No        | Yes
| JavaScript Object Notation  | `.json`    | jsonlite package    | No        | Yes

[arrow]: https://arrow.apache.org/

A **tabular** dataset is one that's structured as a table, with rows and
columns. This reader focuses on tabular datasets, since they're common in
practice and present the fewest programming challenges. Here's an example of a
tabular dataset:

| Fruit  | Quantity | Price
| :----  | -------: | ----:
| apple  | 32       | 1.49
| banana | 541      | 0.79
| pear   | 10       | 1.99

A **text file** is a file that contains human-readable lines of text. You can
check this by opening the file with a text editor such as Microsoft Notepad or
macOS TextEdit. Many file formats use text in order to make the format easier
to work with.

For instance, a **comma-separated values** (CSV) file records a tabular data
using one line per row, with commas separating columns. If you store the table
above in a CSV file and open the file in a text editor, here's what you'll see:

```{.none}
Fruit,Quantity,Price
apple,32,1.49
banana,541,0.79
pear,10,1.99
```

A **binary file** is one that's not human-readable. You can't just read off the
data if you open a binary file in a text editor, but they have a number of
other advantages. Compared to text files, binary files are often faster to read
and take up less storage space (bytes).

As an example, R's built-in binary format is called RDS (which may stand for "R
data serialized"). RDS files are extremely useful for backing up work, since
they can store any kind of R object, even ones that are not tabular. You can
learn more about how to create an RDS file on the `?saveRDS` help page, and how
to read one on the `?readRDS` help page.


## Dataset: CA Least Terns {#sec-ca-least-terns}

The California least tern is a endangered subspecies of seabird that nests
along the coast of California and Mexico. The California Department of Fish and
Wildlife (CDFW) monitors least tern nesting sites across the state to estimate
breeding pairs, fledglings, and predator activity in each annual breeding
season.

![
  A California least tern. Original photo by [Mark Pavelka, U.S. Fish &
  Wildlife Service][pavelka] ([CC BY 2.0][]).
](/images/ca_least_tern_usfws_pacificsw.jpg){
  height=24em
  fig-alt="A gray bird with a white belly, black head, and orange beak sitting
  on a clutch of eggs."
}

[pavelka]: https://www.flickr.com/photos/usfws_pacificsw/5704890596/
[CC BY 2.0]: https://creativecommons.org/licenses/by/2.0/

The CDFW publishes most of the data it collects to the [California Open Data
portal][data.ca.gov]. The examples in this and subsequent chapters use a
cleaned 2000-2023 version of the California least tern data.

[data.ca.gov]: https://data.ca.gov/

:::{.callout-important}
[Click here][ca-least-tern] to download the 2000-2023 California least tern
dataset.

[ca-least-tern]: https://ucdavis.box.com/s/m2w5l2ebp2rey2do5lnn38y1e3u31pin

If you haven't already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/` subdirectory. Download and save
the California least tern dataset in the `data/` subdirectory.
:::

:::{.callout-note
  title="Documentation for 2000-2023 California Least Tern Dataset"
  collapse=true
}

Each row in the dataset contains measurements from one year-site combination.

| Column                | Description
| --------------------: | :----------
| `year`                | Year of the breeding season
| `site_name`           | Site name
| `site_name_2013_2018` | Site name from 2013-2018
| `site_name_1988_2001` | Site name from 1988-2001
| `site_abbr`           | Abbreviated site name
| `region_3`            | Region of state: S.F. Bay, Central, or Southern (includes Ventura)
| `region_4`            | Region of state: S.F. Bay, Central, Ventura, or Southern
| `event`               | Climate events
| `bp_min`              | Reported minimum breeding pairs
| `bp_max`              | Reported maximum breeding pairs
| `fl_min`              | Reported minimum fledges
| `fl_max`              | Reported maximum fledges
| `total_nests`         | Total reported nests (maximum if a range was reported)
| `nonpred_eggs`        | Total non-predator-related mortalities of eggs
| `nonpred_chicks`      | Total non-predator-related mortalities of chicks
| `nonpred_fl`          | Total non-predator-related mortalities of fledges
| `nonpred_ad`          | Total non-predator-related mortalities of adults
| `pred_control`        | Site predator control (yes/no)
| `pred_eggs`           | Total predator-related mortalities of eggs
| `pred_chicks`         | Total predator-related mortalities of chicks
| `pred_fl`             | Total predator-related mortalities of fledges
| `pred_ad`             | Total predator-related mortalities of adults
| `pred_pefa`           | Predation by peregrine falcons (yes/no)
| `pred_coy_fox`        | Predation by coyotes or foxes (yes/no)
| `pred_meso`           | Predation by other mesocarnivores: dogs, cats, skunks, opossums, raccoons, weasels, etc. (yes/no)
| `pred_owlspp`         | Predation by owls (yes/no)
| `pred_corvid`         | Predation by corvids: ravens or crows (yes/no)
| `pred_other_raptor`   | Predation by raptors other than peregrine falcons and owls (yes/no)
| `pred_other_avian`    | Predation by birds other than raptors and corvids (yes/no)
| `pred_misc`           | Predation by other animals (yes/no)
| `total_pefa`          | Total mortalities due to peregrine falcons
| `total_coy_fox`       | Total mortalities due to coyotes and foxes
| `total_meso`          | Total mortalities due to other mesocarnivores
| `total_owlspp`        | Total mortalities due to owls
| `total_corvid`        | Total mortalities due to ravens and crows
| `total_other_raptor`  | Total mortalities due to other raptors
| `total_other_avian`   | Total mortalities due to other birds
| `total_misc`          | Total mortalities due to other animals
| `first_observed`      | Date CA least terns first observed at site
| `last_observed`       | Date CA least terns last observed at site
| `first_nest`          | Date first egg observed at site
| `first_chick`         | Date first chick observed at site
| `first_fledge`        | Date first fledge observed at site

: {.striped .hover tbl-colwidths="[25, 75]"}

The messy source dataset (with more years and more columns) is available
[here][ca-least-tern-source].

[ca-least-tern-source]: https://data.ca.gov/dataset/california-least-tern-monitoring-sites-generalized-cdfw-ds3146
:::

Let's use R to read the California least tern dataset. The dataset is in a file
called is `2000-2023_ca_least_tern.csv`, which suggests it's a CSV file. The
function to read a CSV file is `read.csv`. The function's first and only
required argument is the path to the CSV file.

In the following code, the path to the California least tern dataset is
`data/2000-2023_ca_least_tern.csv`, but it might be different for you,
depending on R's working directory and where you saved the file. We'll save the
result from the `read.csv` function in a variable called `terns`. We can use
this variable to access the data in subsequent code.

```{r}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

:::{.callout-note}
The variable name `terns` is arbitrary; you can choose something different if
you want. However, in general, it's a good habit to choose variable names that
describe the contents of the variable somehow.
:::

If you tried running the line of code above and got an error message, pay
attention to what the error message says, and remember the strategies to get
help from @sec-getting-help. The most common mistake when reading a file is
incorrectly specifying the path, so first check that you got the path right.

If the code ran without errors, it's a good idea to check that the dataset
looks like what the documentation describes. When working with a new dataset,
it usually isn't a good idea to print the whole thing (at least until you know
how big it is). Large datasets can take a long time to print, and the output
can be difficult to read.

Instead, use the `head` function to print only the beginning, or head, of the
dataset:

```{r}
#| eval: false
head(terns)
```

```{r, echo = FALSE}
knitr::kable(head(terns))
```

If you run this code and see a similar table, then congratulations, you've read
your first dataset into R! ✨

The California least terns dataset is tabular---as you might have already
guessed, since it came from a CSV file. In R, it's represented by a **data
frame**, a table with rows and columns. R uses data frames to represent most
(but not all) kinds of tabular data. The `read.csv` function, which you used to
read this data, always returns a data frame.

Typically, each row in a data frame corresponds to a single subject and
is called an **observation**. Each column corresponds to a measurement of the
subject and is called a **feature** or **covariate**.

:::{.callout-note}
Sometimes people also refer to columns as “variables," but we'll try to avoid
this, because in programming contexts a variable is a name for a value (which
might not be a column).
:::

When you first read an object into R, you might not know whether it's a data
frame. One way to check is visually, by printing it (as you just did with
`head`). A better way to check is with the `class` function, which returns
information about what an object is. For a data frame, the result will always
contain `data.frame`:

```{r}
class(terns)
```

You'll learn more about classes in @sec-data-types-classes, but for now you can
use this function to identify data frames.


## Inspecting a Data Frame {#sec-inspecting-data-frame}

Similar to how the `head` function shows the first six rows of a data frame,
the `tail` function shows the last six:

```{r}
tail(terns)
```

If there are lots of columns or the columns are wide, as is the case here, R
wraps the output across lines.

:::{.callout-tip}
Both `head` and `tail` accept an optional second argument that specifies the
number of rows to print:

```{r}
head(terns, 1)
```
:::

One way to get a quick idea of what your data looks like without having to skim
through all the columns and rows is by inspecting its **dimensions**. This is
the number of rows and columns in a data frame, and you can access this
information with the `dim` function:

```{r}
dim(terns)
```

So this dataset has `r nrow(terns)` rows and `r ncol(terns)` columns. As an
alternative to the `dim` function, you can use the `nrow` and `ncol` functions
to get just the number of rows and number of columns, respectively.

Since the columns have names, you might also want to get just these. You can do
that with the `names` or `colnames` functions. Both return the same result:

```{r}
names(terns)
colnames(terns)
```

If the rows have names, you can get those with the `rownames` function. For
this particular dataset, the rows don't have names.


### Summarizing Data

An efficient way to get a sense of what's actually in a dataset is to have R
compute summary information. This works especially well for data frames, but
also applies to other data. R provides two different functions to get
summaries: `str` and `summary`.

The `str` function returns a **structural summary** of an object. This kind of
summary tells us about the structure of the data---the number of rows, the
number and names of columns, what kind of data is in each column, and some
sample values. Here's the structural summary for the least terns dataset:

```{r}
str(terns)
```

This summary lists information about each column, and includes most of what you
found earlier by using several different functions separately. The summary uses
`chr` to indicate columns of text ("characters") and `int` to indicate columns
of integers.

In contrast to `str`, the `summary` function returns a **statistical summary**
of an object. This summary includes summary statistics for each column,
choosing appropriate statistics based on the kind of data in the column. For
numbers, this is generally the mean, median, and quantiles. For categories,
this is the frequencies. Other kinds of statistics are shown for other kinds of
data. Here's the statistical summary for the least terns dataset:

```{r}
summary(terns)
```


### Selecting Columns {#sec-working-with-files-selecting-columns}

You can select an individual column from a data frame by name with `$`, the
dollar sign operator. The syntax is:

```{r}
#| eval: false
VARIABLE$COLUMN_NAME
```

For example, for the least terns dataset, `terns$year` selects the `year`
column, which is the year of observation:

```{r}
terns$year
```

R provides a variety of functions to compute on columns (and other vectors of
data). For instance, what if you want to know the time period the dataset
covers? You can use the `range` function to compute the minimum and maximum of
a column:

```{r}
range(terns$year)
```

So the oldest observations are from `r min(terns$year)` and the newest are from
`r max(terns$year)`, although this function and output doesn't tell us whether
there are observations for the years in between.

You can count the observations for each year with the `table` function:

```{r}
table(terns$year)
```

The `table` function is great for summarizing columns of categories, where
numerical statistics like means and standard deviations aren't defined.

On the other hand, numerical statistics work well for summarizing columns of
numbers. You can use the `mean` function to compute the mean of a column. For
instance, let's compute the mean of the `total_nests` column, which is the
total number of nests seen at a site:

```{r}
mean(terns$total_nests)
```

The result is `NA` because column is missing some values (we'll explain this in
detail in @sec-missing-values). To compute the mean with only the values that
are present, set `na.rm = TRUE` in the call to `mean`:

```{r}
mean(terns$total_nests, na.rm = TRUE)
```

You can also use the dollar sign operator to assign values to columns. For
instance, to assign `2000` to the entire `year` column:

```{r}
#| eval: false
terns$year = 2000
```

Be careful when you do this, as there is no undo. Fortunately, you haven't
saved this change to the least terns dataset to your computer's hard drive yet,
so you can reload the dataset to reset it:

```{r}
#| eval: false
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

In @sec-indexing, you'll learn how to select rows and individual elements from
a data frame, as well as other ways to select columns.


## Exercises


### Exercise

1. Choose a directory on your computer that you're familiar with, such as one
   you created. Determine the path to the directory, then use `list.files` to
   display its contents. Do the files displayed match what you see in your
   system's file browser?

2. What does the `all.files` parameter of `list.files` do? Give an example.


### Exercise

The `read.table` function is another function for reading tabular data. Take a
look at the help file for `read.table`. Recall that `read.csv` reads tabular
data where the values are separated by commas, and `read.delim` reads tabular
data where the values are separated by tabs.

1. What value-separator does `read.table` expect by default?
2. Is it possible to use `read.table` to read a CSV? Explain. If your answer is
   yes, show how to use `read.table` to load the least terns dataset from
   @sec-ca-least-terns.
