Automating Tasks
================

By now, we've learned all of the basic skills necessary to explore a data set
in R. This final chapter is a collection of different topics relevant to
cleaning data sets before and during analysis. Some of them, like loops, also
have other applications.


Conditional Expressions
-----------------------

Sometimes you'll need code to do different things, depending on a condition.
_If-statements_ provide a way to write conditional code.

For example, suppose we want to greet one person differently from the others:
```{r}
name = "Nick"
if (name == "Nick") {
   # If name is Nick:
   message("We went down the TRUE branch")
   msg = "Hi Nick, nice to see you again!"
} else {
   # Anything else:
   msg = "Nice to meet you!"
}
```

Indent code inside of the if-statement by 2 or 4 spaces. Indentation makes your
code easier to read.

The condition in an if-statement has to be a scalar:
```{r, error = TRUE}
name = c("Nick", "Susan")
if (name == "Nick") {
   msg = "Hi Nick!"
} else {
   msg = "Nice to meet you!"
}
```

You can chain together if-statements:
```{r}
name = "Susan"
if (name == "Nick") {
   msg = "Hi Nick, nice to see you again!"
} else if (name == "Peter") {
   msg = "Go away Peter, I'm busy!"
} else {
   msg = "Nice to meet you!"
}
msg
```

If-statements return the value of the last expression in the evaluated block:
```{r}
name = "Tom"
msg = if (name == "Nick") {
   "Hi Nick, nice to see you again!"
} else {
   "Nice to meet you!"
}
msg
```

Curly braces `{ }` are optional for single-line expressions:
```{r}
name = "Nick"
msg = if (name == "Nick") "Hi Nick, nice to see you again!" else
   "Nice to meet you!"
msg
```

But you have to be careful if you don't use them:
```{r, error = TRUE}
# NO GOOD:
msg = if (name == "Nick")
   "Hi Nick, nice to see you again!"
else
   "Nice to meet you!"
```

The `else` block is optional:
```{r}
msg = "Hi"
name = "Tom"
if (name == "Nick")
   msg = "Hi Nick, nice to see you again!"
msg
```

When there's no `else` block, the value of the `else` block is `NULL`:
```{r}
name = "Tom"
msg = if (name == "Nick")
   "Hi Nick, nice to see you again!"
msg
```


Functions
---------

The main way to interact with R is by calling functions, which was first
explained way back in Section \@ref(calling-functions). Since then, you've
learned how to use many of R's built-in functions. This section explains how
you can write your own functions.

To start, let's briefly review what functions are, and some of the jargon
associated with them. It's useful to think of functions as factories: raw
materials (inputs) go in, products (outputs) come out. We can also represent
this visually:

```
         +-------+
-- in -->|   f   |-- out -->
         +-------+
```

Programmers use several specific terms to describe the parts and usage of
functions:

* _Parameters_ are placeholder variables for inputs.
    + _Arguments_ are the actual values assigned to the parameters in a call.
* The _return value_ is the output.
* The _body_ is the code inside.
* _Calling_ a function means using a function to compute something.

Almost every command in R is a function, even the arithmetic operators and the
parentheses! You can view the body of a function by typing its name without
trailing parentheses (in contrast to how you call functions). The body of a
function is usually surrounded by curly braces `{}`, although they're optional
if the body only contains one line of code. Indenting code inside of curly
braces by 2-4 spaces also helps make it visually distinct from other code.

For example, let's look at the body of the `append` function, which appends a
value to the end of a list or vector:

```{r}
append
```

Don't worry if you can't understand everything the `append` function's code
does yet. It will make more sense later on, after you've written a few
functions of your own.

Many of R's built-in functions are not entirely written in R code. You can spot
these by calls to the special `.Primitive` or `.Internal` functions in their
code.

For instance, the `sum` function is not written in R code:

```{r}
sum
```

The `function` keyword creates a new function. Here's the syntax:

```
function(parameter1, parameter2, ...) {
  # Your code goes here
  
  # The result goes here
}
```

A function can have any number of parameters, and will automatically return the
value of the last line of its body.

A function is a value, and like any other value, if you want to reuse it, you
need to assign it to variable. Choosing descriptive variable names is a good
habit. For functions, that means choosing a name that describes what the
function does. It often makes sense to use verbs in function names.

Let's create a function that detects negative numbers. It should take a vector
of numbers as input, compare them to zero, and then return the logical result
from the comparison as output. Here's the code to do that:

```{r}
is_negative = function(x) {
  x < 0
}

# Since this is a one-line function, the curly braces are optional:
is_negative = function(x) x < 0
```
The name of the function, `is_negative`, describes what the function does and
includes a verb. The parameter `x` is the input. The return value is the result
of `x < 0`, since we put that expression on the last line. 

Any time you write a function, the first thing you should do afterwards is test
that it actually works. Let's try the `is_negative` function on a few test
cases:

```{r}
x = c(5, -1, -2, 0, 3)

is_negative(6)
is_negative(-1.1)
is_negative(x)
```

Notice that the parameter `x` inside the function is different from the
variable `x` we created outside the function. We'll learn more about this in
Section \@ref(variable-scope-lookup), but for now just remember that parameters
and variables inside of a function are separate from variables outside of a
function.

A _default argument_ is an argument assigned to a parameter if no argument is
assigned in the call to the function. You can use `=` to assign default
arguments to parameters when you define a function with the `function` keyword.


For example, let's write a function that gets the largest values in a vector.
We'll make the number of values to get a parameter, with a default argument of
`5`. Here's the code and some test cases:

```{r}
top = function(x, n = 5) {
  sorted = sort(x, decreasing = TRUE)
  head(sorted, n)
}

y = c(-6, 7, 10, 3, 1, 15, -2)
top(y, 3)
top(y)
```

### Returning Values

We've already seen that a function will automatically return the value of its
last line.

The `return` keyword causes a function to return a result immediately, without
running any subsequent code in its body. It only makes sense to use `return`
from inside of an if-statement. If your function doesn't have any
if-statements, you don't need to use `return`.

For example, suppose we want the `is_negative` function to return `NA` if the
argument isn't a number. This is an ideal case to use `return`, since we skip
the computation `x < 0` when the argument isn't a number. Here's the new
version of the function, as well as a new test case:

```{r}
is_negative = function(x) {
  if (!is.numeric(x))
    return (NA)
  
  # TRUE for negative numbers, FALSE otherwise
  x < 0
}

is_negative("hi")
```

It's idiomatic to only use `return` when strictly necessary. 

A function returns one R object, but sometimes computations have multiple
results. In that case, return the results in a vector, list, or other data
structure.

For example, let's make a function that computes the mean and median for a
vector. We'll return the results in a named list, although we could also use a
named vector:

```{r}
compute_mean_med = function(x) {
  m1 = mean(x)
  m2 = median(x)
  list(mean = m1, median = m2)
}
compute_mean_med(c(1, 2, 3, 1))
```

The names make the result easier to understand for the caller of the function,
although they certainly aren't required here.


### Planning Your Functions

Before you write a function, it's useful to go through several steps:

1. Write down what you want to do, in detail. It can also help to
   draw a picture of what needs to happen.

2. Check whether there's already a built-in function. Search online and in the
   R documentation.

3. Write the code to handle a simple case first. For data science
   problems, use a small dataset at this step.

Let's apply this in one final example: a function that detects leap years. A
year is a leap year if either of these conditions is true:

* It is divisible by 4 and not 100
* It is divisible by 400

That means the years 2004 and 2000 are leap years, but the year 2200 is not.
Here's the code and a few test cases:

```{r}
# If year is divisible by 4 and not 100 -> leap
# If year is divisible by 400 -> leap
year = 2004
is_leap = function(year) {
  if (year %% 4 == 0 & year %% 100 != 0) {
    leap = TRUE
  } else if (year %% 400 == 0) {
    leap = TRUE
  } else {
    leap = FALSE
  }
  leap
}
is_leap(400)
is_leap(1997)
```

Functions are the building blocks for solving larger problems. Take a
divide-and-conquer approach, breaking large problems into smaller steps. Use a
short function for each step. This approach makes it easier to:

* Test that each step works correctly.
* Modify, reuse, or repurpose a step.


Loops
-----

One major benefit of using a programming language like R is that repetitive
tasks can be automated. We've already seen two ways to do this:

1. Vectorization, introduced in Section \@ref(vectorization)

2. Apply functions, introduced in Section \@ref(apply-functions)

Both of these are _iteration strategies_. They _iterate_ over some object, and
compute something for each element. Each one of these computations is one
_iteration_. Vectorization is the most efficient iteration strategy, but only
works with vectorized functions and vectors. Apply functions are more
flexible---they work with any function and any data structure with
elements---but less efficient and less concise.

A _loop_ is another iteration strategy, one that's even more flexible than
apply functions. Besides being flexible, loops are a feature of almost all
modern programming languages, so it's useful to understand them. In R, there
are two kinds of loops. We'll learn both.

### For-loops

A _for-loop_ runs a block of code once for each element of a vector or list.
The `for` keyword creates a for-loop. Here's the syntax:

```
for (I in DATA) {
  # Your code goes here
}
```

The variable `I` is called the _induction variable_. At the beginning of each
iteration, `I` is assigned the next element of the vector or list `DATA`. The
loop iterates once for each element of `DATA`, unless you use a keyword to exit
the loop early (more about this in Section \@ref(break-next)). As with
if-statements and functions, the curly braces `{ }` are only required if the
body contains multiple lines of code.

Unlike the other iteration strategies, loops do not automatically return a
result. You have complete control over the output, which means that anything
you want to save must be assigned to a variable.

For example, let's make a loop that repeatedly adds a number to a running total
and squares the new total. We'll use a variable `total` to keep track of the
running total as the loop iterates:

```{r}
numbers = c(-1, 1, -3, 2)

total = 0
for (number in numbers) {
  total = (total + number)^2
}

total
```

Use for-loops when some or all of the iterations depend on results from other
iterations. If the iterations are not dependent, use one of:

1. Vectorization (because it's faster)
2. Apply functions (because they're idiomatic)

In some cases, you can use vectorization even when the iterations are
dependent. For example, you can use vectorization to compute the sum of the
cubes of several numbers:

```{r}
sum(numbers^3)
```

### While-loops

A _while-loop_ runs a block of code repeatedly as long as some condition is
`TRUE`. The `while` keyword creates a while-loop. Here's the syntax:

```
while (CONDITION) {
  # Your code goes here
}
```

The `CONDITION` should be a scalar logical value or an expression that returns
one. At the beginning of each iteration, `CONDITION` is checked, and the loop
exits if it is `FALSE`. As always, the curly braces `{ }` are only required if
the body contains multiple lines of code.

For example, suppose you want to add up numbers from 0 to 50, but stop as soon
as the total is greater than 50:

```{r}
num50 = seq(0, 50)

total = 0
i = 1
while (total < 50) {
  total = total + num50[i]
  message("i is ", i, " total is ", total)
  i = i + 1
}

total
i
```

While-loops are a generalization of for-loops. They tend to be most useful when
you don't know how many iterations will be necessary. For example, suppose you
want to repeat a computation until the result falls within some range of
values.


### Saving Multiple Results

Loops often produce a different result for each iteration. If you want to save
more than one result, there are a few things you must do.

First, set up an index vector. The index vector should usually be congruent to
the number of iterations or the input. The `seq_along` function returns a
congruent index vector when passed a vector or list. For instance, let's make
in index for the `numbers` vector from Section \@ref(for-loops):

```{r}
index = seq_along(numbers)
```

The loop will iterate over the index rather than the input, so the induction
variable will track the current iteration number. On the first iteration, the
induction variable will be 1, on the second it will be 2, and so on. Then you
can use the induction variable and indexing to get the input for each
iteration.

Second, set up an empty output vector or list. This should usually be congruent
to the input, or one element longer (the extra element comes from the initial
value). R has several functions for creating vectors. We've already seen a few,
but here are more:

* `logical`, `integer`, `numeric`, `complex`, and `character` to create an
  empty vector with a specific type and length

* `vector` to create an empty vector with a specific type and length

* `rep` to create a vector by repeating elements of some other vector

Empty vectors are filled with `FALSE`, `0`, or `""`, depending on the type of
the vector. Here are some examples:

```{r}
logical(3)
numeric(4)
rep(c(1, 2), 2)
```

Let's create an empty numeric vector congruent to `numbers`:

```{r}
n = length(numbers)
result = numeric(n)
```

As with the input, you can use the induction variable and indexing to set the
output for each iteration.

Creating a vector or list in advance to store something, as we've just done, is
called _preallocation_. Preallocation is extremely important for efficiency in
loops. Avoid the temptation to use `c` or `append` to build up the output bit
by bit in each iteration.

Finally, write the loop, making sure to get the input and set the output.
Here's the loop for the squared sums example:

```{r}
for (i in index) {
  prev = if (i > 1) result[i - 1] else 0
  result[i] = (numbers[i] + prev)^2
}
result
```


### Break & Next

The `break` keyword causes a loop to immediately exit. It only makes sense to
use `break` inside of an if-statement.

For example, suppose we want to print each string in a vector, but stop at the
first missing value. We can do this with `break`:

```{r}
my_messages = c("Hi", "Hello", NA, "Goodbye")

for (msg in my_messages) {
  if (is.na(msg))
    break

  message(msg)
}
```

The `next` keyword causes a loop to immediately go to the next iteration. As
with `break`, it only makes sense to use `next` inside of an if-statement.

Let's modify the previous example so that missing values are skipped, but don't
cause printing to stop. Here's the code:

```{r}
for (msg in my_messages) {
  if (is.na(msg))
    next

  message(msg)
}
```

These keywords work with both for-loops and while-loops.


### Example: Fibonacci


As a final example, let's use a loop to compute Fibonacci numbers. Each
Fibonacci number is the sum of the previous two. The first few Fibonacci
numbers are:

```
1 1 2 3 5 8 13
```

We can use a loop to compute the first `n` Fibonacci numbers. Make sure to
choose an appropriate index and to preallocate for the output:

```{r}
n = 10
fib = c(1, 1, numeric(n - 2))

for (i in 3:n) {
  fib[i] = fib[i - 1] + fib[i - 2]
}

fib
```

If you want to reuse this code, you could make it a function:

```{r}
fibonacci = function(n = 10) {
  fib = c(1, 1, numeric(n - 2))

  for (i in 3:n) {
    fib[i] = fib[i - 1] + fib[i - 2]
  }

  fib
}

fibonacci(30)
```



Planning for Iteration
----------------------

At first it may seem difficult to decide if and what kind of iteration to use.
Start by thinking about whether you need to do something over and over. If you
don't, then you probably don't need to use iteration. If you do, then try
iteration strategies in this order:

1. vectorization
2. apply functions
    * Try an apply function if iterations are independent.
3. for/while-loops
    * Try a for-loop if some iterations depend on others.
    * Try a while-loop if the number of iterations is unknown.
4. recursion (which isn't covered here)
    * Convenient for naturally recursive problems (like Fibonacci),
      but often there are faster solutions.


Start by writing the code for just one iteration. Make sure that code works;
it's easy to test code for one iteration.

When you have one iteration working, then try using the code with an iteration
strategy (you will have to make some small changes). If it doesn't work, try to
figure out which iteration is causing the problem. One way to do this is to use
`message` to print out information. Then try to write the code for the broken
iteration, get that iteration working, and repeat this whole process.


Exercises
---------

