Appendix
========

More About Comparisons
----------------------


### Equality

The `==` operator is the primary way to test whether two values are equal, as
explained in Section \@ref(comparisons). Nonetheless, equality can be defined
in many different ways, especially when dealing with computers. As a result, R
also provides several different functions to test for different kinds of
equality. This describes tests of equality in more detail, and also describes
some other important details of comparisons.


#### The `==` Operator

The `==` operator tests whether its two arguments have the exact same
representation as a _[binary number][binary]_ in your computer's memory. Before
testing the arguments, the operator applies R's rules for vectorization
(Section \@ref(vectorization)), recycling (Section \@ref(recycling)), and
implicit coercion (Section \@ref(implicit-coercion)). Until you've fully
internalized these three rules, some results from the equality operator  may
seem surprising. For example:

```{r}
# Recycling:
c(1, 2) == c(1, 2, 1, 2)

# Implicit coercion:
TRUE == 1
TRUE == "TRUE"
1 == "TRUE"
```

The length of the result from the equality operator is usually the same as its
longest argument (with some exceptions).

[binary]: https://en.wikipedia.org/wiki/Binary_number


#### The `all.equal` Function

The `all.equal` function tests whether its two arguments are equal up to some
acceptable difference called a _tolerance_. Computer representations for
decimal numbers are inherently imprecise, so it's necessary to allow for very
small differences between computed numbers. For example:

```{r}
x = 0.5 - 0.3
y = 0.3 - 0.1

# FALSE on most machines:
x == y

# TRUE:
all.equal(x, y)
```

The `all.equal` function does not apply R's rules for vectorization, recycling,
or implicit coercion. The function returns `TRUE` when the arguments are equal,
and returns a string summarizing the differences when they are not. For
instance:

```{r}
all.equal(1, c(1, 2, 1))
```

The `all.equal` function is often used together with the `isTRUE` function,
which tests whether the result is `TRUE`:

```{r}
all.equal(3, 4)
isTRUE(all.equal(3, 4))
```

You should generally use the `all.equal` function when you want to compare
decimal numbers.


#### The `identical` Function

The `identical` function checks whether its arguments are completely identical,
including their metadata (names, dimensions, and so on). For instance:

```{r}
x = list(a = 1)
y = list(a = 1)
z = list(1)

identical(x, y)
identical(x, z)
```

The `identical` function does not apply R's rules for vectorization, recycling,
or implicit coercion. The result is always a single logical value.

You'll generally use the `identical` function to compare non-vector objects
such as lists or data frames. The function also works for vectors, but most of
the time the equality operator `==` is sufficient.


### The `%in%` Operator

Another common comparison is to check whether elements of one vector are
_contained_ in another vector at any position. For instance, suppose you want
to check whether `1` or `2` appear anywhere in a longer vector `x`. Here's how
to do it:

```{r}
x = c(3, 4, 2, 7, 3, 7)
c(1, 2) %in% x
```

R returns `FALSE` for the `1` because there's no `1` in `x`, and returns `TRUE`
for the `2` because there is a `2` in `x`.

Notice that this is different from comparing with the equality operator `==`.
If you use use the equality operator, the shorter vector is recycled until its
length matches the longer one, and then compared element-by-element. For the
example, this means only the elements at odd-numbered positions are compared to
`1`, and only the elements at even-numbered positions are compared to `2`:

```{r}
c(1, 2) == x
```


### Summarizing Comparisons

The comparison operators are vectorized, so they compare their arguments
element-by-element:

```{r}
c(1, 2, 3) < c(1, 3, -3)
c("he", "saw", "her") == c("she", "saw", "him")
```

What if you want to summarize whether all the elements in a vector are equal
(or unequal)? You can use the `all` function on any logical vector to get a
summary. The `all` function takes a vector of logical values and returns `TRUE`
if all of them are `TRUE`, and returns `FALSE` otherwise:

```{r}
all(c(1, 2, 3) < c(1, 3, -3))
```

The related `any` function returns `TRUE` if any one element is `TRUE`, and
returns `FALSE` otherwise:

```{r}
any(c("hi", "hello") == c("hi", "bye"))
```

  
### Other Pitfalls

New programmers sometimes incorrectly think they need to append `== TRUE` to
their comparisons. This is redundant, makes your code harder to understand, and
wastes computational time. Comparisons already return logical values. If the
result of the comparison is `TRUE`, then `TRUE == TRUE` is again just `TRUE`.
If the result is `FALSE`, then `FALSE == TRUE` is again just `FALSE`. Likewise,
if you want to invert a condition, choose an appropriate operator rather than
appending `== FALSE`.



Variable Scope & Lookup
-----------------------

### Local Variables

A variable's *scope* is the section of code where it exists and is accessible.
The `exists` function checks whether a variable is in scope:

```{r}
exists("zz")
zz = 3
exists("zz")
```

When you create a function, you create a new scope. Variables defined inside of
a function are *local* to the function. Local variables cannot be accessed from
outside:

```{r, error = TRUE}
rescale = function(x, center, scale) {
  centered = x - center
  centered / scale
}

centered
exists("centered")
```

Local variables are reset each time the function is called:

```{r}
f = function() {
  is_z_in_scope = exists("z")
  z = 42
  
  is_z_in_scope
}

f()
f()
```


### Lexical Scoping

A function can use variables defined outside (non-local), but only if those
variables are in scope where the function was **defined**. This property is
called *lexical scoping*.

Let's see how this works in practice. First, we'll define a variable `cats` and
then define a function `get_cats` in the same place (the top level, not inside
any functions). As a result, the `cats` variable is in scope inside of the
`get_cats` function:

```{r}
cats = 3
get_cats = function() cats
get_cats()
```

Now let's define a variable `dogs` inside of a function `create_dogs`. We'll
also define a function `get_dogs` at the top level. The variable `dogs` is not
in scope at the top level, so it's not in scope inside of the `get_dogs`
function:

```{r, error = TRUE}
create_dogs = function() {
  dogs = "hello"
}
get_dogs = function() dogs
create_dogs()
get_dogs()
```

Variables defined directly in the R console are *global* and available to any
function.

Local variables *mask* (hide) non-local variables with the same name:

```{r}
get_parrot = function() {
  parrot = 3
  
  parrot
}
parrot = 42
get_parrot()
```

There's one exception to this rule. We often use variables that refer to
functions in calls:

```{r}
#mean()
```

In this case, the variable must refer to a function, so R ignores local
variables that aren't functions. For example:

```{r}
my_mean = function() {
  mean = 0
  
  mean(c(1, 2, 3))
}
my_mean()
my_get_cats = function() {
  get_cats = 10
  
  get_cats()
}
my_get_cats()
```


### Dynamic Lookup

Variable lookup happens when a function is **called**, not when it's defined.
This is called *dynamic lookup*.

For example, the result from `get_cats`, which accesses the global variable
`cats`, changes if we change the value of `cats`:

```{r}
cats = 10
get_cats()
cats = 20
get_cats()
```

### Summary

This section covered a lot of details about R's rules for variable scope and
lookup. Here are the key takeaways:

* Function definitions (or `local()`) create a new scope.

* Local variables
    + Are private
    + Get reset for each call
    + Mask non-local variables (exception: function calls)

* *Lexical scoping*: where a function is **defined** determines which non-local
  variables are in scope.

* *Dynamic lookup*: when a function is **called** determines values of
  non-local variables.


String Processing
-----------------

So far, we've mostly worked with numbers or categories that are ready to use
for data analysis. In practice, data sets often require some cleaning before or
during data analysis. One common data cleaning task is editing or extracting
parts of strings.

We'll use the stringr package to process strings. Like ggplot2 (Section
\@ref(data-visualization)), the package is part of the [Tidyverse][tidy]. R
also has built-in functions for string processing. The main advantage of
stringr is that its functions use a common set of parameters, so they're easier
to learn and remember.

stringr has detailed [documentation][stringr-docs] and also a
[cheatsheet][stringr-cheat-sheet].

[stringr-docs]: https://stringr.tidyverse.org/
[stringr-cheat-sheet]: https://github.com/rstudio/cheatsheets/blob/master/strings.pdf
[tidy]: https://www.tidyverse.org/

The first time you use stringr, you'll have to install it with
`install.packages` (the same as any other package). Then you can load the
package with the `library` function:

```{r}
# install.packages("stringr")
library("stringr")
```

The typical syntax of a stringr function is:

```
str_NAME(string, pattern, ...)
```

Where:

* `NAME` describes what the function does
* `string` is the string to search within or transform
* `pattern` is the pattern to search for
* `...` is additional, function-specific arguments

The `str_detect` function detects whether the pattern appears within the
string. Here's an example:

```{r}
str_detect("hello", "el")

str_detect("hello", "ol")
```

Most of the stringr functions are vectorized in the `string` parameter. For
instance:

```{r}
str_detect(c("hello", "goodbye", "lo"), "lo")
```

Most of the stringr functions also have support for [_regular
expressions_][regex], a powerful language for describing  patterns. Several
punctuation characters, such as `.` and `?` have special meanings in the
regular expressions language. You can disable these special meanings by putting
the pattern in a call to `fixed`:

```{r}
str_detect("a", ".")
str_detect("a", fixed("."))
```

You can learn more about regular expressions [here][regex-r4ds].

[regex]: https://en.wikipedia.org/wiki/Regular_expression
[regex-r4ds]: https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions


There are a lot of stringr functions. We'll focus on two that are especially
important, and some of their variants:

* `str_split`
* `str_replace`
 
You can find a complete list of stringr functions, with examples, in the
documentation.

### Splitting Strings

The `str_split` function splits the string at each position that matches the
pattern. The characters that match are thrown away.

For example, suppose we want to split a sentence into words. Since there's a
space between each word, we can use a space as the pattern:
```{r}
x = "The students in this workshop are great!"

result = str_split(x, " ")
result
```

The `str_split` function always returns a list with one element for each input
string. Here the list only has one element because `x` only has one element. We
can get the first element with:
```{r}
result[[1]]
```

We have to use the extraction operator `[[` here because `x` is a list (for a
vector, we could use the indexing operator `[` instead). Notice that in the
printout for `result`, R gives us a hint that we should use `[[` by printing
`[[1]]`.

To see why the function returns a list, consider what happens if we try to
split two different sentences at once:

```{r}
x = c(x, "Are you listening?")

result = str_split(x, " ")
result[[1]]
result[[2]]
```

Each sentence has a different number of words, so the vectors in the result
have different lengths. So a list is the only way to store both.

The `str_split_fixed` function is almost the same as `str_split`, but takes a
third argument for the maximum number of splits to make. Because the number of
splits is fixed, the function can return the result in a matrix instead of a
list. For example:

```{r}
str_split_fixed(x, " ", 3)
```

The `str_split_fixed` function is often more convenient than `str_split`
because the `n`th piece of each input string is just the `n`th column of the
result.

For example, suppose we want to get the area code from some phone numbers:

```{r}
phones = c("717-555-3421", "629-555-8902", "903-555-6781")
result = str_split_fixed(phones, "-", 3)

result[, 1]
```


### Replacing Parts of Strings

The `str_replace` function replaces the pattern the first time it appears in
the string. The replacement goes in the third argument.

For instance, suppose we want to change the word `"dog"` to `"cat"`:
```{r}
x = c("dogs are great, dogs are fun", "dogs are fluffy")
str_replace(x, "dog", "cat")
```


The `str_replace_all` function replaces the pattern every time it appears in
the string:
```{r}
str_replace_all(x, "dog", "cat")
```

We can also use the `str_replace` and `str_replace_all` functions to delete
part of a string by setting the replacement to the empty string `""`.

For example, suppose we want to delete the comma:
```{r}
str_replace(x, ",", "")
```

In general, stringr functions with the `_all` suffix affect all matches.
Functions without `_all` only affect the first match.



Date Processing
---------------

Besides strings, dates and times are another kind of data that require special
attention to prepare for analysis. This is especially important if you want to
do anything that involves sorting dates, like making a line plot with dates on
one axis. Dates may not be sorted correctly if they haven't been converted to
one of R's date classes.

There several built-in functions and also many packages for date processing. As
with visualization and string processing, the Tidyverse packages have the best
combination of simple design and clear documentation. There are three Tidyverse
packages for processing dates and times:

* [lubridate][], the primary package for working with dates and times
* [hms][], a package specifically for working with times
* [clock][], a new package for working with dates and times

[lubridate]: https://lubridate.tidyverse.org/
[hms]: https://hms.tidyverse.org/
[clock]: https://clock.r-lib.org/

We'll focus on the lubridate package. As always, you'll have to install the
package if you haven't already, and then load it:

```{r}
# install.packages("lubridate")
library("lubridate")
```

The most common task is to convert a string into a date or time class. For
instance, when you load a data set, you might have dates that look like this:

```{r}
dates = c("Jan 10, 2021", "Sep 3, 2018", "Feb 28, 1982")
dates
```

These are strings, so it's relatively difficult to sort the dates, do
arithmetic on them, or extract just one part (such as the year). There are
several lubridate functions to automatically convert strings into dates. They
are named with one letter for each part of the date. For instance, the dates in
the example have the month (m), then the day (d), and then the year (y), so we
can use the `mdy` function:

```{r}
result = mdy(dates)
result
class(result)
```

Notice that the dates now have class `Date`, one of R's built-in classes for
representing dates, and that they print differently. You can find a full list
of the automatic string to date conversion functions in the lubridate
documentation.

Occasionally, a date string may have a format that lubridate can't convert
automatically. In that case, you can use the `fast_strptime` function to
describe the format in detail. At a minimum, the function requires two
arguments: the vector of strings to convert and a format string.

The format string describes the format of the dates, and is based on the syntax
of `strptime`, a function provided by many programming languages for converting
strings to dates (including R). In a format string, a percent sign `%` followed
by a character is called a _specification_ and has a special meaning. Here are
a few of the most useful ones:

| Specification | Description      | January 29, 2015
|:------------- |:---------------- |:----------------
| `%Y`          | 4-digit year     | 2015
| `%y`          | 2-digit year     | 15
| `%m`          | 2-digit month    | 01
| `%B`          | full month name  | January
| `%b`          | short month name | Jan
| `%d`          | day of month     | 29
| `%%`          | literal %        | %

You can find a complete list in `?fast_strptime`. Other characters in the
format string do not have any special meaning. Write the format string so that
it matches the format of the dates you want to convert.

For example, let's try converting an unusual time format:
```{r}
odd_time = "6 minutes, 32 seconds after 10 o'clock"
fast_strptime(odd_time, "%M minutes, %S seconds after %H o'clock")
```

R usually represents dates with the class `Date`, and date-times with the
classes `POSIXct` and `POSIXlt`. The difference between the two date-time
classes is somewhat technical, but you can read more about it in `?POSIXlt`.

There is no built-in class to represent times alone, which is why the result in
the example above includes a date. Nonetheless, the hms package provides the
`hms` class to represent times without dates.

Once you've converted a string to a date, the lubridate package provides a
variety of functions to get or set the parts individually. Here are a few
examples:

```{r}
day(result)
month(result)
```

You can find a complete list in the lubridate documentation.
